---
title: "Code chunks"
author: "Boyoung"
date: "2018년 8월 7일"
output: 
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

## Code chunks

To run code inside an R Markdown document, you need to insert a chunk. There are three ways to do so:

1. The keyboard shortcut Cmd/Ctrl + Alt + I

1. The ??Insert?? button icon in the editor toolbar.

1. By manually typing the chunk delimiters ```{r} and ```.

R code chunks can be used as a means render R output into documents or to simply display code for illustration. Here is a simple R code chunk that will result in both the code and it??s output being included:

```{r}
summary(cars)
```

To display the output of a code chunk but not the underlying R code, you specify the echo=FALSE option:

```{r, echo=FALSE}
summary(cars)
```

Note that R code chunks can also be used to render plots. To display a plot while omitting the code used to generate the plot you??d do this:

```{r, echo=FALSE}
plot(cars)
```

To display R code without evaluating it, you specify the eval=FALSE chunk option:

```{r, eval=FALSE}
summary(cars)
```
